GrapeCPNet: A Self-supervised Point Cloud Completion Network for 3D Phenotyping of Grape Bunches

### 2025-05-08 response2reviewer.tex

\manuscript{
    the branching structure could potentially be captured during early stages of thinning, as suggested by \citep{du_instance_2023}.
    \added[id=R1]{Lastly, with the development of lidar scanning techniques, some companies also provide low-cost products to improve 3D scanning results in the field, such as XGrids (XGRIDS LIMITED, \url{https://www.xgrids.com}).}
    Such data could enable studies on growth prediction for optimizing harvesting times and assessments of branching structure deviations within crowded berry bunches.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Regarding the completion net, I would have liked to see ablation studies here as I believe this is the main  contribution of this work. For example, how does the number of cuts  influence the results of the completion metrics? How does the number of  output points in the centroid-based contour influence the snowflake net?
\end{reviewercomment}

\response{
    We thank the reviewer for the insightful suggestions regarding ablation studies for our completion network. 
    We would like to clarify that GrapeCPNet is not an entirely new network architecture, but rather a combination of features from two existing models (PoinTr and SnowFlakeNet). 
    This approach differs from traditional network design methodologies. 
    Therefore, performing the suggested ablation studies may not be practical or aligned with our main research objectives.

    However, when treating each network component as an independent module, our results already include comparative experiments between individual networks and the combined architecture. 
    As shown in Table 3, we compared the performance of: (1) PoinTr alone; (2) SnowFlakeNet alone; (3) Our combined GrapeCPNet. 
    The results demonstrate that the combined architecture outperforms either individual network.

    To support future research in this area, we will make our source code and training data publicly available. 
    This will allow other researchers to examine how preprocessing parameters may affect the network's performance.

    We have also added the above discussion in the results section.
}

\manuscript{
    this paper further validated the efficacy of the 3D phenotypic traits of grape by the proposed method.

    \added[id=R1]{
        Since GrapeCPNet combines features from two existing models (PoinTr and SnowFlakeNet) rather than introducing a completely new architecture,
        ablation studies may not align with our research focus.
        However, we conducted comparative experiments by treating each network as an independent module.
        As shown in Table~\ref{tbl:4}, we evaluated three approaches: (1) PoinTr alone, (2) SnowFlakeNet alone, and (3) our combined GrapeCPNet.
        The results show that GrapeCPNet achieves better performance than either individual network.
        To facilitate future research, we will release our source code and training data at \url{https://github.com/I3-Laboratory/GrapeCPNet}.
        This will enable interested researchers to examine how preprocessing parameters influence the network's performance.
    }

    \textbf{3.3 Phenotypic Traits Calculation Results}
}

\manuscript{
    \textbf{Data availability}

    The \added[id=R1]{demo dataset, source codes} and model weight for the \replaced[id=R1]{academic}{test} usage is available \replaced[id=R1]{at}{here:} \url{https://github.com/I3-Laboratory/GrapeCPNet}. 
    The whole dataset \added[id=R1]{and codes} used in this paper is available upon request.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Finally, is there a reason why the approach of Blok et al. 2025 is not being used as a  baseline? I would have liked to see a non-learning-based baseline as  well, like Marangoz et al. 2022
\end{reviewercomment}

\response{
    We appreciate the reviewer's suggestion regarding additional baselines.

    The comparison with \citet{blok_highthroughput_2025} was not included initially because these were independent parallel research projects during our manuscript preparation phase. 
    Following editorial recommendations during the submission process, we subsequently added discussions of comparable works in the revised manuscript, including it into the discussion part after \citet{blok_highthroughput_2025} been accepted.

    Additionally, the network architectures target substantially different objectives. 
    This grape completion involves partial completion using the same sensor type, while the potato network focuses on fusing different sensors with an emphasis on processing speed and real-time performance. 
    Specifically, \citet{blok_highthroughput_2025} works with incomplete point clouds from RGB-D sensors and complete point clouds from SfM, which exhibit significant differences in shape and texture. 
    Our approach aims for better completion performance on ellipsoid berry shapes with smooth surfaces, whereas \citet{blok_highthroughput_2025} focuses on irregular and elongated potato tuber shapes. 
    Although both networks perform completion tasks, their purposes and input data differ. 
    For the future work, it may be interesting to do a comparation of two networks on inverted crops.

    It is also the similar reason why we not comparing with \citet{marangoz_fruit_2022}.
    The work of \citet{marangoz_fruit_2022} generates cube-like point clouds for peppers, which would may not be suitable for ellipsoid grape berries with smooth surfaces.
    Additionally, re-implementation their algorithm is time-consuming and unlikely to benefit our current grape project.
    Our ellipsoid surface fitting is a similar non-learning-based approach, serving for berry traits analysis in this study (section~\ref{sec:ellipsoid}). 
    Therefore, non-learning-based approach may not be appropriate as a baseline comparison

    We have also added the above discussion in the introduction and discussion section.
}

\manuscript{
    provided an appropriate symmetrical plane can be identified.
    \citet{marangoz_fruit_2022} proposed an super-ellipsoids matching approach to \replaced[id=R1]{generate cube-like point clouds for peppers}{ map fruits on plants} and estimate their shape.
    However, these methods are limited to specific fruit characteristics and require manual design of algorithms according to the features of the fruit
}

\manuscript{
    training datasets of complete and incomplete point clouds.
    In our \replaced[id=R1]{recent}{previous} work on potato tuber completion networks \citep{blok_highthroughput_2025}, a distinguishable pin was 
}

\manuscript{
    these automatically generated paired training datasets were successfully used to train several completion models (Table~\ref{tbl:4}).

    \added[id=R1]{
        Although our recent work proposed a completion network for partial potato tubers \citep{blok_highthroughput_2025}, the purposes and input data types differ. 
        This grape completion study uses the same sensor type, whereas the potato network focuses on fusing different sensors with an emphasis on processing speed and real-time performance on fast-moving conveyors. 
        Specifically, the tuber completion study processes incomplete point clouds from RGB-D sensors and complete point clouds from SfM, which show significant differences in shape and texture. 
        Additionally, the grape completion targets smooth ellipsoid berry surfaces, while potato tubers have more irregular and elongated shapes. 
        For future work, it may be interesting to compare the two networks on inverted crops.
    }
    We \replaced[id=R1]{also}{have} found that some researchers
}

### 2025-05-08 main.tex
R1 finish

Another challenge after the success of instance berry segmentation is the incomplete representation of berry surfaces. 
The aggregated cluster growth causes the berries to squeezed against each other, leading to self occlusion and resulting in an incomplete point cloud of berry surfaces \citep{du_instance_2023}. 
Therefore, point cloud instance segmentation can only capture partial information rather than that of whole berries.
To address this issue and consequently an underestimation of the berry traits, point cloud shape completion is one of the most commonly used solutions.
Point cloud completion is the task of generating a complete 3D shape given partial point cloud input \citep{tesema_point_2024}. 
For instance, \citet{ge_symmetrybased_2020} assume that strawberries are symmetrical, suggesting that their shape can be completed by mirroring the existing shape, provided an appropriate symmetrical plane can be identified.
\citet{marangoz_fruit_2022} proposed an super-ellipsoids matching approach to map fruits on plants and estimate their shape.
However, these methods are limited to specific fruit characteristics and require manual design of algorithms according to the features of the fruit, lacking the potential for generalized application.
Utilizing deep learning methods, researchers have designed point cloud completion networks \citep{wang_unsupervised_2021}. 
\citet{park_deepsdf_2019} introduced DeepSDF to achieve the completion of object point cloud models. It uses a learned continuous signed distance function to represent a class of shapes in latent space and is tolerant to noisy 3D input data. 
\citet{magistri_contrastive_2022} utilized DeepSDF and its latent space in an encoder-decoder network that converted a single RGB-D frame to a 3D shape of a complete fruit. 
This encoder-decoder network was extended and optimized by \citet{blok_highthroughput_2025} for high-throughput 3D shape completion of potato tubers on a commercial harvester. 
For grape berry application, the shape of the berry can be roughly approached by the deformation of a sphere or ellipsoid depending on the species. 
Existing datasets from other fruits cannot be applied directly, and manually creating paired training datasets is time-consuming and labor-intensive.
Therefore, there is a demand to develop an analysis pipeline, capable of completing the detailed surface of an incomplete berry point cloud while requiring less labor-intensive training data preparation.

@@

\subsection{Phenotypic Traits Calculation and Validation}

@@

@@

For berry completion using deep learning, the most challenging and time-consuming task is preparing the paired training datasets of complete and incomplete point clouds.
In our previous work on potato tuber completion networks \citep{blok_highthroughput_2025}, a distinguishable pin was used as a reference point to align the complete tuber scans obtained via photogrammetry with the incomplete tuber scans captured using an RGB-D camera.
A pin-guided point cloud matching algorithm was then applied to calculate the transformation matrix for pairing the datasets.
Potato tubers, being single objects with clear textures and shape variation, presented fewer challenges when aligning data collected by two different types of sensors.
However, grape berries pose a significantly more difficult problem due to their highly rounded shapes, weak texture features, and strong reflectance, which complicates the creation of paired data.
As a result, the previous pin-guided point cloud pairing method used for potato tuber datasets is not effective for grape berries.
Thus, we simulated random sphere cutting on the complete berry point clouds to generate thousands of paired incomplete point clouds, as described in Subsection~\ref{sec:berrycut}.
As demonstrated in Subsection~\ref{sec:32}, these automatically generated paired training datasets were successfully used to train several completion models (Table~\ref{tbl:4}).
We have found that some researchers \citep{pan_panoptic_2023,magistri_efficient_2024} have improved the completion effect by encoding a prior knowledge of the geometry of the object into the weights of the neural network. 
Considering the agricultural plant completion task, our proposed GrapeCPNet, which combines centroid-based sparse completion with diffusion-based dense completion, outperformed other networks in terms of performance.
We found that GrapeCPNet performed a little bit worse in the case of strange berry shapes, such as drop-shaped \added[id=R2]{berries, particularly when berries were sereverly occluded}.
We subsequently consider utilizing the prior knowledge of berry geometry to achieve more detailed morphological feature extraction and representation to improve the quality of completion.

@@

\section*{Data availability}
The dataset and model weight for the test usage is available here: https://github.com/I3-Laboratory/GrapeCPNet. The whole dataset used in this paper is available upon request.

>>>>

Another challenge after the success of instance berry segmentation is the incomplete representation of berry surfaces. 
The aggregated cluster growth causes the berries to squeezed against each other, leading to self occlusion and resulting in an incomplete point cloud of berry surfaces \citep{du_instance_2023}. 
Therefore, point cloud instance segmentation can only capture partial information rather than that of whole berries.
To address this issue and consequently an underestimation of the berry traits, point cloud shape completion is one of the most commonly used solutions.
Point cloud completion is the task of generating a complete 3D shape given partial point cloud input \citep{tesema_point_2024}. 
For instance, \citet{ge_symmetrybased_2020} assume that strawberries are symmetrical, suggesting that their shape can be completed by mirroring the existing shape, provided an appropriate symmetrical plane can be identified.
\citet{marangoz_fruit_2022} proposed an super-ellipsoids matching approach to \replaced[id=R1]{generate cube-like point clouds for peppers}{ map fruits on plants} and estimate their shape.
However, these methods are limited to specific fruit characteristics and require manual design of algorithms according to the features of the fruit, lacking the potential for generalized application.
Utilizing deep learning methods, researchers have designed point cloud completion networks \citep{wang_unsupervised_2021}. 
\citet{park_deepsdf_2019} introduced DeepSDF to achieve the completion of object point cloud models. It uses a learned continuous signed distance function to represent a class of shapes in latent space and is tolerant to noisy 3D input data. 
\citet{magistri_contrastive_2022} utilized DeepSDF and its latent space in an encoder-decoder network that converted a single RGB-D frame to a 3D shape of a complete fruit. 
This encoder-decoder network was extended and optimized by \citet{blok_highthroughput_2025} for high-throughput 3D shape completion of potato tubers on a commercial harvester. 
For grape berry application, the shape of the berry can be roughly approached by the deformation of a sphere or ellipsoid depending on the species. 
Existing datasets from other fruits cannot be applied directly, and manually creating paired training datasets is time-consuming and labor-intensive.
Therefore, there is a demand to develop an analysis pipeline, capable of completing the detailed surface of an incomplete berry point cloud while requiring less labor-intensive training data preparation.

@@

\subsection{Phenotypic Traits Calculation and Validation}
\label{sec:ellipsoid}

@@

\added[id=R1]{
    Since GrapeCPNet combines features from two existing models (PoinTr and SnowFlakeNet) rather than introducing a completely new architecture,
    ablation studies may not align with our research focus.
    However, we conducted comparative experiments by treating each network as an independent module.
    As shown in Table~\ref{tbl:4}, we evaluated three approaches: (1) PoinTr alone, (2) SnowFlakeNet alone, and (3) our combined GrapeCPNet.
    The results show that GrapeCPNet achieves better performance than either individual network.
    To facilitate future research, we will release our source code and training data at \url{https://github.com/I3-Laboratory/GrapeCPNet}.
    This will enable interested researchers to examine how preprocessing parameters influence the network's performance.
}

@@

For berry completion using deep learning, the most challenging and time-consuming task is preparing the paired training datasets of complete and incomplete point clouds.
In our \replaced[id=R1]{recent}{previous} work on potato tuber completion networks \citep{blok_highthroughput_2025}, a distinguishable pin was used as a reference point to align the complete tuber scans obtained via photogrammetry with the incomplete tuber scans captured using an RGB-D camera.
A pin-guided point cloud matching algorithm was then applied to calculate the transformation matrix for pairing the datasets.
Potato tubers, being single objects with clear textures and shape variation, presented fewer challenges when aligning data collected by two different types of sensors.
However, grape berries pose a significantly more difficult problem due to their highly rounded shapes, weak texture features, and strong reflectance, which complicates the creation of paired data.
As a result, the previous pin-guided point cloud pairing method used for potato tuber datasets is not effective for grape berries.
Thus, we simulated random sphere cutting on the complete berry point clouds to generate thousands of paired incomplete point clouds, as described in Subsection~\ref{sec:berrycut}.
As demonstrated in Subsection~\ref{sec:32}, these automatically generated paired training datasets were successfully used to train several completion models (Table~\ref{tbl:4}).

\added[id=R1]{
    Although our recent work proposed a completion network for partial potato tubers \citep{blok_highthroughput_2025}, the purposes and input data types differ. 
    This grape completion study uses the same sensor type, whereas the potato network focuses on fusing different sensors with an emphasis on processing speed and real-time performance on fast-moving conveyors. 
    Specifically, the tuber completion study processes incomplete point clouds from RGB-D sensors and complete point clouds from SfM, which show significant differences in shape and texture. 
    Additionally, the grape completion targets smooth ellipsoid berry surfaces, while potato tubers have more irregular and elongated shapes. 
    For future work, it may be interesting to compare the two networks on inverted crops.
}
We \replaced[id=R1]{also}{have} found that some researchers \citep{pan_panoptic_2023,magistri_efficient_2024} have improved the completion effect by encoding a prior knowledge of the geometry of the object into the weights of the neural network. 
Considering the agricultural plant completion task, our proposed GrapeCPNet, which combines centroid-based sparse completion with diffusion-based dense completion, outperformed other networks in terms of performance.
We found that GrapeCPNet performed a little bit worse in the case of strange berry shapes, such as drop-shaped \added[id=R2]{berries, particularly when berries were sereverly occluded}.
We subsequently consider utilizing the prior knowledge of berry geometry to achieve more detailed morphological feature extraction and representation to improve the quality of completion.

@@

\section*{Data availability}
The \added[id=R1]{demo dataset, source codes} and model weight for the \replaced[id=R1]{academic}{test} usage is available \replaced[id=R1]{at}{here:} \url{https://github.com/I3-Laboratory/GrapeCPNet}. 
The whole dataset \added[id=R1]{and codes} used in this paper is available upon request.
